data_collation

### 2024-05-03 scripts/data_collation.Rmd
almost finished organizing swaath data

# use rbind.fill() method in R to combine data frames with different columns
swath_data <- rbind.fill(swath, swath_2018, swath_count, FtWeatherill_swath_2016)

@@

# load species names and species code dataframe
species.metadata <- read.csv("data/species_code_metadata.csv")

# Convert species names to codes
for (col in names(swath)) {
  swath[, col] <- species.metadata$code[match(swath[, col], species.metadata$species.substrate)]
}

#some of the species in swath dataframe are common names and species metadata in species names

>>>>

# Load species names and species code data frame
species.metadata <- read.csv("data/species_code_metadata.csv")

# Function to rename columns using metadata
# Get the unique column names from the swath dataframe
column_names <- colnames(swath)

# Create a lookup table with name-to-code mappings
name_to_code <- setNames(species.metadata$code, species.metadata$name)

# Rename the columns using the lookup table
  new_column_names <- sapply(column_names, function(name) {
    if (name %in% names(name_to_code)) {
      return(name_to_code[name])
    } else {
      return(name)
    }
  })

# Rename the columns
colnames(swath) <- new_column_names

# Convert data in numeric columns to numeric
swath.con <- swath %>%
  mutate(across(AGH:BLOA, as.numeric))

# Melt only columns AGH:WHEL into long format
swath_long <- swath.con %>%
  pivot_longer(cols = AGH:BLOA, names_to = "SP_CODE", values_to = "value")

# Need to now move values to binned swatch categories as columns
# Pivot the dataframe into wide format
swath_binned <- swath_long %>%
  pivot_wider(names_from = swath, values_from = value)

# make columns lowercase
# Get the current column names
current_names <- colnames(swath_binned)

# Convert the column names to lowercase while preserving the first letter
new_names <- sapply(current_names[1:6], function(x) {
  paste0(toupper(substring(x, 1, 1)), tolower(substring(x, 2)))
})

# Assign the new column names to the first six columns of the dataframe
colnames(swath_binned)[1:6] <- new_names

# rename swatch category columns in swath count and swath 208 dataframe
swath_2018 <- swath_2018 %>%
  rename(
    "0-20 IN" = "X0.20.IN",
    "20-40 IN" = "X20.40.IN",
    "40-20 OFF" = "X40.20.OFF",
    "20-0 OFF" = "X20.0.OFF"
  )

swath_count <- swath_count %>%
  rename(
    "0-20 IN" = "X0.20.IN",
    "20-40 IN" = "X20.40.IN",
    "40-20 OFF" = "X40.20.OFF",
    "20-0 OFF" = "X20.0.OFF"
  )

# Remove four columns in a specific order swath_2018
swath_2018 <- swath_2018 %>%
  select(-12:-15)

# Remove four columns in a specific order swath_count
swath_count <- swath_count %>%
  select(-12:-15)

# Remove four columns in a specific order FtWeatherill_swath_2016
FtWeatherill_swath_2016 <- FtWeatherill_swath_2016 %>%
  select(-9:-14)

# make columns lowercase
# Get the current column names
current_names <- colnames(FtWeatherill_swath_2016)

# Convert the column names to lowercase while preserving the first letter
new_names <- sapply(current_names[1:6], function(x) {
  paste0(toupper(substring(x, 1, 1)), tolower(substring(x, 2)))
})

# Assign the new column names to the first six columns of the dataframe
colnames(FtWeatherill_swath_2016)[1:6] <- new_names

# use rbind.fill() method in R to combine data frames with different columns
swath_data <- rbind.fill(swath_binned, swath_2018, swath_count, FtWeatherill_swath_2016)

@@

#still need to edit ft.wetherill dataframe
